---
title: "Ensemble learning"
author: "Dr. Shirin Glander"
date: "April 18, 2017"
output: html_document
---

https://www.analyticsvidhya.com/blog/2017/02/introduction-to-ensembling-along-with-implementation-in-r/





Previous research in data mining has devised numerous different algorithms for learning tasks. While an individual algorithm might already work decently, one can usually obtain a better predictive by combining several. This approach is referred to as ensemble learning.
Common examples include random forests, boosting and AdaBost in particular.

I would like to give a basic overview of ensemble learning. Ensemble learning involves combining multiple predictions derived by different techniques in order to create a stronger overall prediction. For example, the predictions of a random forest, a support vector machine, and a simple linear model may be combined to create a stronger final prediction set. The key to creating a powerful ensemble is model diversity. An ensemble with two techniques that are very similar in nature will perform more poorly than a more diverse model set.

Some ensemble learning techniques, such as Bayesian model combination and stacking, attempt to weight the models prior to combining them.

Random Forests are an ensemble learning method for classification
and regression
I It combines multiple individual decision trees by means of bagging
I Overcomes the problem of overfitting decision trees

Create many decision trees by bagging
2 Inject randomness into decision trees
a. Tree grows to maximum size and is left unpruned
I Deliberate overfitting: i. e. each tree is a good model on its own
b. Each split is based on randomly selected subset of attributes
I Reduces correlation between different trees
I Otherwise, many trees would just select the very strong predictors
3 Ensemble trees (i. e. the random forest) vote on categories by majority

<br>

## The data

https://archive.ics.uci.edu/ml/datasets/Mammographic+Mass

Attribute Information:
   1. BI-RADS assessment: 1 to 5 (ordinal)  
   2. Age: patient's age in years (integer)
   3. Shape: mass shape: round=1 oval=2 lobular=3 irregular=4 (nominal)
   4. Margin: mass margin: circumscribed=1 microlobulated=2 obscured=3 ill-defined=4 spiculated=5 (nominal)
   5. Density: mass density high=1 iso=2 low=3 fat-containing=4 (ordinal)
   6. Severity: benign=0 or malignant=1 (binominal)

```{r message=FALSE, warning=FALSE}
library(tidyverse)

data <- read.table("mammographic_masses.data.txt", sep = ",") %>%
  mutate(V2 = as.numeric(as.character(V2)),
         V6 = as.factor(V6))

data[data == "?"] <- NA

# Remove NA observations
data <- na.omit(data)
```

```{r mds_plot, echo=FALSE, eval=FALSE}
# Multidimensional Scaling 
select(data, -6) %>%
  dist() %>%
  cmdscale %>%
  as.data.frame() %>%
  mutate(group = data$V6) %>%
  ggplot(aes(x = V1, y = V2, color = group)) +
    geom_point()
```

```{r fig.width=5, fig.height=3}
data %>%
  ggplot(aes(x = V6)) +
    geom_bar()
```

## 

```{r warning=FALSE, message=FALSE}
# configure multicore
library(doParallel)
cl <- makeCluster(detectCores())
registerDoParallel(cl)

library(caret)
```

```{r}
set.seed(42)
index <- createDataPartition(data$V6, p = 0.7, list = FALSE)
train_data <- data[index, ]
test_data  <- data[-index, ]
```

```{r}
models <- c("rf", "xgbTree", "knn")
```

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=4, eval=FALSE}
for (model in models) {
  set.seed(42)
  model_train <- caret::train(V6 ~ .,
                           data = train_data,
                           method = model,
                           trControl = trainControl(method = "repeatedcv", 
                                                    number = 10, 
                                                    repeats = 5,
                                                    verboseIter = FALSE)) 
  assign(paste0("model_", model), model_train)
}
```

```{r echo=FALSE, eval=FALSE}
save(model_rf, file = "model_rf.RData")
save(model_xgbTree, file = "model_xgbTree.RData")
save(model_knn, file = "model_knn.RData")
```

```{r echo=FALSE}
for (model in models) {
  load(paste0("model_", model, ".RData"))
}
```

```{r}
predictions <- data.frame()
for (model in models) {
  prediction <- data.frame(id = 1:nrow(test_data),
                           model = model,
                           predict(get(paste0("model_", model)), test_data, type = "prob"),
                           actual = test_data$V6)
  prediction$prediction <- gsub("X", "", colnames(prediction)[3:4][apply(prediction[, 3:4], 1, which.max)])
  prediction$correct <- ifelse(prediction$actual == prediction$prediction, "correct", "wrong")
  predictions <- rbind(predictions, prediction)
}
```

```{r}
ensemble <- predictions %>%
  group_by(id) %>%
  dplyr::summarise(X0 = sum(X0) / 3,
                   X1 = sum(X1) / 3) %>%
  mutate(actual = test_data$V6)

ensemble$prediction <- gsub("X", "", colnames(ensemble)[2:3][apply(ensemble[, 2:3], 1, which.max)])
ensemble$correct <- ifelse(ensemble$actual == ensemble$prediction, "correct", "wrong")
```

```{r}
predictions %>%
  group_by(model) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)

ensemble %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)
```

```{r}
ensemble2 <- predictions %>%
  mutate(X0 = ifelse(model == "rf", X0 * 2, ifelse(model == "xgbTree", X0 * 2, X0)),
         X1 = ifelse(model == "rf", X1 * 2, ifelse(model == "xgbTree", X1 * 2, X0))) %>%
  group_by(id) %>%
  dplyr::summarise(X0 = sum(X0) / 3,
                   X1 = sum(X1) / 3) %>%
  mutate(actual = test_data$V6)

ensemble2$prediction <- gsub("X", "", colnames(ensemble2)[2:3][apply(ensemble2[, 2:3], 1, which.max)])
ensemble2$correct <- ifelse(ensemble2$actual == ensemble2$prediction, "correct", "wrong")


ensemble2 %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)
```

```{r}
predictions %>%
  group_by(model) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)

ensemble %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)

```

```{r}
predictions %>%
  group_by(model, actual) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)

predictions %>%
  group_by(model, prediction) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)
```


```{r}
ensemble %>%
  group_by(actual) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)

ensemble %>%
  group_by(prediction) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n)
```

## Bayesian model combination

## Model stacking

## Boosting & AdaBoost

Bagging: Bootstrap Aggregation
I Meta strategy design to accuracy of machine learning algorithms
I Improvements for unstable procedures
→ Neural networks, trees and linear regression with subset selection,
rule learning (opposed to k-NN, linear regression, SVM)
I Idea: Reuse the same training algorithm several times on different
subsets of the training data
I When classifier needs random initialization (e. g. k-means), very these
across each run

Combine multiple classifiers to improve classification accuracy
I Works together with many different types of classifiers
I None of the classifier needs extremely good, only better than chance

Idea: train classifiers on a subset of the training data that is most
informative given the current classifiers

High-level algorithm
1 Fit a simple model to a subsample of the data
2 Identify misclassified observations, i. e. that are hard to predict
3 Focus subsequent learners on these samples and get them right
4 Combine weak learners to form a complex predictor

```{r}
library(mboost)
```

AdaBoosting
Instead of resampling, reweight misclassified training examples


```{r}
library(ada)

```

---

If you are interested in more machine learning posts, check out [the category listing for **machine_learning** on my blog](https://shiring.github.io/categories.html#machine_learning-ref).

---

```{r}
sessionInfo()
```

