---
title: "Roadtrip 2016"
author: "Dr. Shirin Glander"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

# Mapping GPS data from our USA/ Canada Roadtrip

This September we went on a roadtrip to the US and Canada. We flew to Chicago, went west across Michigan to Ann Arbor, Detroit, Toronto, then south again to Pittsburgh, from where we followed the Ohio river and headed to Louisville. From there, we went to Paducah, KY and back north again to Chicago.

Of course, we had our trusty GPS to guide us along the way, the data from which I downloaded and used to play around with.

## Loading the data

I used the **XML** package to load the gpx files following [this blog post](https://www.r-bloggers.com/stay-on-track-plotting-gps-tracks-with-r/).

```{r echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
library(XML)

myfiles <- list.files(path = "../gpx/gpx_files", full.names = TRUE)

for (i in 1:length(myfiles)){

  # One of the files seems to be broken, but since the file is one with earlier data
  # not from the trip, I will simply skip it
  tryCatch({
    pfile <- htmlTreeParse(myfiles[i], useInternalNodes = T)

  }, error=function(e){cat("ERROR\n")})

  if (exists("pfile")){
    # Get all elevations, times and coordinates via the respective xpath
    elevations <- as.numeric(as.character(xpathSApply(pfile, path = "//trkpt/ele", xmlValue)))
    times <- xpathSApply(pfile, path = "//trkpt/time", xmlValue)
    coords <- xpathSApply(pfile, path = "//trkpt", xmlAttrs)
    # Extract latitude and longitude from the coordinates
    lats <- as.numeric(as.character(coords["lat",]))
    lons <- as.numeric(as.character(coords["lon",]))
    # Put everything in a dataframe and get rid of old variables
    geodf <- data.frame(lat = lats, lon = lons, ele = elevations, time = times)

    if (i == 1){
      geodata <- geodf
      } else {
      geodata <- rbind(geodata, geodf)
    }
  }
  rm(pfile)
}
geodata_all <- geodata
```

I then did some cleaning of the date and time column and removed all the old track that were not part of our trip.

```{r echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
# Transforming the time column
geodata_all$time <- strptime(geodata_all$time, format = "%Y-%m-%dT%H:%M:%SZ")

# ordering by date
library(plyr)
geodata_all <- arrange(geodata_all, time)

# removing all tracks from Germany
geodata <- geodata_all[which(as.numeric(as.character(geodata_all$lon)) < 0),]

# adding a column with day for separating the plotting by day lateron
geodata$day <- as.factor(format(as.Date(geodata$time,format="%Y-%m-%d"), "%d"))
```

We have `r nrow(geodata)` data points with a range of: `r range(as.character(geodata$time))`.

```{r echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
head(geodata)
summary(geodata)
str(geodata)
```

## Get additional data from Google Maps

Somehow the GPS didn't record the last two days of our trip back to Chicago, so I added some of the way points with the Google Maps Geocoding and Directions APIs (accessed via the **googleway** package).

```{r echo=TRUE, eval=FALSE}
library(googleway)

# on th 26th we had lunch at cracker barrel
cracker_barrel <- google_geocode(address = "2101 N Kenyon Rd, Urbana, IL 61802, USA",
                                 key = "your_key",
                                 language = "en",
                                 simplify = TRUE)

cracker_barrel_data <- data.frame(lat = cracker_barrel$results$geometry$location$lat,
                                  lon = cracker_barrel$results$geometry$location$lng,
                                  ele = "NA", time = as.POSIXlt("2016-09-26 19:00:00"))

# then we stayed the night in Aurora/ Naperville
super8 <- google_geocode(address = "4228 Longmeadow Dr, Aurora, IL 60504, USA",
                         key = "your_key",
                         language = "en",
                         simplify = TRUE)

super8_data <- data.frame(lat = super8$results$geometry$location$lat,
                          lon = super8$results$geometry$location$lng,
                          ele = "NA", time = as.POSIXlt("2016-09-26 23:30:00"))

# on the 27th we went back to the airport to fly home
ohare <- google_geocode(address = "10000 W O'Hare Ave, Chicago, IL 60666, USA",
                        key = "your_key",
                        language = "en",
                        simplify = TRUE)

ohare_data <- data.frame(lat = ohare$results$geometry$location$lat,
                         lon = ohare$results$geometry$location$lng,
                         ele = "NA", time = as.POSIXlt("2016-09-27 18:00:00"))
```

Now I want the travel information between these points to get more waypoints along the route.

```{r echo=TRUE, eval=FALSE}
## Route from Paducah to Cracker Barrel

route_pad_cracker <- google_directions(origin = c(geodata$lat[nrow(geodata)], geodata$lon[nrow(geodata)]),
                  destination = c(cracker_barrel_data$lat, cracker_barrel_data$lon),
                  mode = "driving",
                  key = "your_key",
                  language = "en",
                  simplify = TRUE)

route_pad_cracker_legs <- as.data.frame(route_pad_cracker$routes$legs)
route_pad_cracker_legs_steps <- as.data.frame(route_pad_cracker_legs$steps)

## Route from Cracker Barrel to Aurora

route_cracker_aurora <- google_directions(origin = c(cracker_barrel_data$lat, cracker_barrel_data$lon),
                                          destination = c(super8_data$lat, super8_data$lon),
                                          mode = "driving",
                                          key = "your_key",
                                          language = "en",
                                          simplify = TRUE)

route_cracker_aurora_legs <- as.data.frame(route_cracker_aurora$routes$legs)
route_cracker_aurora_legs_steps <- as.data.frame(route_cracker_aurora_legs$steps)

## Route from Cracker Aurora to O'Hare

route_aurora_ohare <- google_directions(origin = c(super8_data$lat, super8_data$lon),
                                        destination = c(ohare_data$lat, ohare_data$lon),
                                        mode = "driving",
                                        key = "your_key",
                                        language = "en",
                                        simplify = TRUE)

route_aurora_ohare_legs <- as.data.frame(route_aurora_ohare$routes$legs)
route_aurora_ohare_legs_steps <- as.data.frame(route_aurora_ohare_legs$steps)

# Combining the data sets
geodata_2 <- data.frame(lat = c(geodata$lat,
                                route_pad_cracker_legs$start_location$lat, route_pad_cracker_legs_steps$start_location$lat,
                                route_pad_cracker_legs$end_location$lat, cracker_barrel_data$lat,
                                route_cracker_aurora_legs$start_location$lat, route_cracker_aurora_legs_steps$start_location$lat,
                                route_cracker_aurora_legs$end_location$lat, super8_data$lat,
                                route_aurora_ohare_legs$start_location$lat, route_aurora_ohare_legs_steps$start_location$lat,
                                route_aurora_ohare_legs$end_location$lat, ohare_data$lat),
                      lon = c(geodata$lon, route_pad_cracker_legs$start_location$lng, route_pad_cracker_legs_steps$start_location$lng,
                              route_pad_cracker_legs$end_location$lng, cracker_barrel_data$lon,
                              route_cracker_aurora_legs$start_location$lng, route_cracker_aurora_legs_steps$start_location$lng,
                              route_cracker_aurora_legs$end_location$lng, super8_data$lon,
                              route_aurora_ohare_legs$start_location$lng, route_aurora_ohare_legs_steps$start_location$lng,
                              route_aurora_ohare_legs$end_location$lng, ohare_data$lon),
                      day = c(as.character(geodata$day), rep("26",
                                               length(c(route_pad_cracker_legs$start_location$lat,
                                                        route_pad_cracker_legs_steps$start_location$lat,
                                                        route_pad_cracker_legs$end_location$lat, cracker_barrel_data$lat,
                                                        route_cracker_aurora_legs$start_location$lat,
                                                        route_cracker_aurora_legs_steps$start_location$lat,
                                                        route_cracker_aurora_legs$end_location$lat, super8_data$lat))),
                              rep("27",
                                  length(c(route_aurora_ohare_legs$start_location$lat, route_aurora_ohare_legs_steps$start_location$lat,
                                           route_aurora_ohare_legs$end_location$lat, ohare_data$lat)))))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
geodata_2 <- read.table("../gpx/geodata_plus_maps.txt", header = T, sep = "\t")
```

## Plotting

I tried several packages for plotting but ended up with **ggmap** as the nicest looking solution.

### Overview of the trip

Each color shows a different day

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height= 10}
library(ggmap)

bc_bbox <- make_bbox(lat = lat, lon = lon, data = geodata_2, f = .10)
map <- get_map(location = bc_bbox, maptype = "roadmap")

ggmap(map, darken = c(0.3, "white")) + geom_point(aes(x = lon, y = lat, col = factor(day), group = factor(day)), data = geodata_2) +
  geom_path(aes(x = lon, y = lat, col = factor(day)), data = geodata_2[which(geodata_2$day %in% c("26", "27")), ]) + 
  theme(legend.position="none") + labs(x="Longitude", y="Latitude", title="Roadtrip 2016")
```

### Road trip with stops

First, I need to obtain the geocodes from the stops (again via Google Maps API).

```{r echo=TRUE, eval=FALSE}
chicago <- google_geocode(address = "Chicago",
                          key = "your_key",
                          language = "en",
                          simplify = TRUE)

stjospeh <- google_geocode(address = "St. Joseph, MI",
                          key = "your_key",
                          language = "en",
                          simplify = TRUE)

annarbor <- google_geocode(address = "Ann Arbor, MI",
                           key = "your_key",
                           language = "en",
                           simplify = TRUE)

detroit <- google_geocode(address = "Detroit, MI",
                           key = "your_key",
                           language = "en",
                           simplify = TRUE)

burlington <- google_geocode(address = "Burlington, ON",
                          key = "your_key",
                          language = "en",
                          simplify = TRUE)

ashtabula <- google_geocode(address = "Ashtabula",
                             key = "your_key",
                             language = "en",
                             simplify = TRUE)

pittsburgh <- google_geocode(address = "Pittsburgh",
                            key = "your_key",
                            language = "en",
                            simplify = TRUE)

marietta <- google_geocode(address = "Marietta, OH",
                             key = "your_key",
                             language = "en",
                             simplify = TRUE)

maysville <- google_geocode(address = "Maysville, KY",
                             key = "your_key",
                             language = "en",
                             simplify = TRUE)

louisville <- google_geocode(address = "Louisville, KY",
                             key = "your_key",
                             language = "en",
                             simplify = TRUE)

paducah <- google_geocode(address = "Paducah, KY",
                             key = "your_key",
                             language = "en",
                             simplify = TRUE)

ann_text <- data.frame(x = c(chicago$results$geometry$location$lng,
                             stjospeh$results$geometry$location$lng,
                             annarbor$results$geometry$location$lng,
                             detroit$results$geometry$location$lng,
                             burlington$results$geometry$location$lng,
                             ashtabula$results$geometry$location$lng,
                             pittsburgh$results$geometry$location$lng,
                             marietta$results$geometry$location$lng,
                             maysville$results$geometry$location$lng,
                             louisville$results$geometry$location$lng[1],
                             paducah$results$geometry$location$lng),
                       y = c(chicago$results$geometry$location$lat,
                             stjospeh$results$geometry$location$lat,
                             annarbor$results$geometry$location$lat,
                             detroit$results$geometry$location$lat+0.1, #adjusting Detroit label to avoid overlapping with Ann Arbor
                             burlington$results$geometry$location$lat,
                             ashtabula$results$geometry$location$lat,
                             pittsburgh$results$geometry$location$lat,
                             marietta$results$geometry$location$lat,
                             maysville$results$geometry$location$lat,
                             louisville$results$geometry$location$lat[1],
                             paducah$results$geometry$location$lat),
                       labs = c(paste0(chicago$results$formatted_address),
                                paste0(stjospeh$results$formatted_address),
                                paste0(annarbor$results$formatted_address),
                                paste0(detroit$results$formatted_address),
                                paste0(burlington$results$formatted_address),
                                paste0(ashtabula$results$formatted_address),
                                paste0(pittsburgh$results$formatted_address),
                                paste0(marietta$results$formatted_address),
                                paste0(maysville$results$formatted_address),
                                paste0(louisville$results$formatted_address[1]),
                                paste0(paducah$results$formatted_address)
                                ))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
ann_text <- read.table("../gpx/ann_text.txt", header = T, sep = "\t")
```

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height= 10}
map <- get_map(location = bc_bbox, maptype = "watercolor")

roadtrip_map <- ggmap(map, darken = c(0.3, "white")) + geom_point(aes(x = lon, y = lat, col = factor(day), group = factor(day)), data = geodata_2, alpha = 0.3) +
  geom_path(aes(x = lon, y = lat, col = factor(day)), data = geodata_2[which(geodata_2$day %in% c("26", "27")), ], alpha = 0.5) +
  theme(legend.position="none") + labs(x="Longitude", y="Latitude", title="Roadtrip 2016")
roadtrip_map + geom_text(data = ann_text, aes(x, y, label = labs), size = 4)
```

### Road trip statistics

#### How many kilometers did we drive?

```{r echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
# removing duplicate entries
geodata_2 <- geodata_2[!duplicated(geodata_2), ]

# Shifting vectors for latitude and longitude to include end position
shift.vec <- function (vec, shift){
  if(length(vec) <= abs(shift)){
    rep(NA ,length(vec))
  } else {
    if (shift >= 0) {
      c(rep(NA, shift), vec[1:(length(vec)-shift)]) }
    else {
      c(vec[(abs(shift)+1):length(vec)], rep(NA, abs(shift)))
    }
  }
}

geodata_2$lat.p1 <- shift.vec(geodata_2$lat, -1)
geodata_2$lon.p1 <- shift.vec(geodata_2$lon, -1)

# Calculating distances between points (in metres) with the function pointDistance from the 'raster' package.

library(raster)
geodata_2$dist.to.prev <- apply(geodata_2, 1, FUN = function (row) {
  pointDistance(c(as.numeric(as.character(row["lat.p1"])),
                  as.numeric(as.character(row["lon.p1"]))),
                c(as.numeric(as.character(row["lat"])), as.numeric(as.character(row["lon"]))),
                lonlat = T) # Parameter 'lonlat' has to be TRUE!
})
```

We drove approximately `r round(sum(as.numeric(as.character(geodata_2$dist.to.prev)), na.rm = TRUE)*0.001, digits = 2)` km in 18 days. Not counting the ~6 days where we stayed in one place and didn't drive much, that's around `r round(sum(as.numeric(as.character(geodata_2$dist.to.prev)), na.rm = TRUE)*0.001/12, digits = 2)` km/ day.

### Identifying local time

The time stamps are in CEST - Central European Summer Time. We arrived in the Central Time Zone (CDT) but spent the majority of our trip in Estern Time Zone (EDT).

#### Converting to CDT and EDT

```{r echo=TRUE, eval=FALSE}
library(lubridate)

geodata$time_CDT <- with_tz(geodata$time, tz="America/Chicago")
geodata$time_EDT <- with_tz(geodata$time, tz="America/New_York")

# extracting day for EDT
geodata$day_EDT <- as.factor(format(as.Date(geodata$time_EDT,format="%Y-%m-%d"), "%d"))
```

#### Identifying the local time based on geocode

```{r echo=TRUE, eval=FALSE}
# the days where we crossed time zones were the 10th and 23rd

library(googleway)

geodata_timezone <- geodata
geodata_timezone$local_time <- "NA"

for (i in 1:nrow(geodata)){
  if (is.na(nchar(strsplit(as.character(i/100), "\\.")[[1]][2]))){
    print(i)
  }

  if (geodata[i, "day_EDT"] == "10"){
    data_row <- geodata[i, ]
    google_timezone_api <- google_timezone(location = c(data_row$lat, data_row$lon),
                                           timestamp = as.POSIXct(data_row$time_EDT), # because the majority of our trip was in EDT
                                           key = "AIzaSyBkqrk8mBqqvao-W-jKkL2VhHCGLWqZVCY",
                                           simplify = TRUE,
                                           language = "en")

    time_zone <- google_timezone_api$timeZoneId

    local_time <- as.POSIXlt(with_tz(data_row$time, tz=paste(time_zone)))

    geodata_timezone$local_time[i] <- as.character(local_time)

    rm(data_row, google_timezone_api, time_zone)

  } else {

    if (geodata[i, "day_EDT"] == "09"){
      geodata_timezone$local_time[i] <- as.character(geodata_timezone$time_CDT[i])
    } else {
      geodata_timezone$local_time[i] <- as.character(geodata_timezone$time_EDT[i])
    }
  }
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
geodata_timezone <- read.table("../gpx/geodata_timezone.txt", header = T, sep = "\t")
```

### Plotting elevation

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height= 10}
theme_set(theme_bw()) # Change the theme to my preference

labels <- c("9" = "9th, Chicago", "10" = "10th, St. Joseph", "11" = "11th, Ann Arbor", "12" = "12th, Detroit", "13" = "13th, Burlington",
            "14" = "14th, Burlington", "15" = "15th, Toronto", "16" = "16th, Niagara & Ashtabula", "17" = "17th, Pittsburgh",
            "18" = "18th, Pittsburgh", "19" = "19th, Marietta", "20" = "20th, Maysville", "21" = "21st, Louisville",
            "22" = "22nd, Louisville", "23" = "23rd, Paducah")

ggplot(aes(x = as.POSIXlt(local_time), y = ele), data = geodata_timezone) +
  geom_point(size = 0.5) + facet_wrap( ~ day_EDT, ncol=3, scales = "free_x", labeller=labeller(day_EDT = labels)) +
  labs(x="Local Time", y="Elevation", title="Elevation per day and time")
```

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=15, fig.height= 5}
day_df <- data.frame(row.names = unique(geodata_timezone$day_EDT),
                     group = unique(geodata_timezone$day_EDT),
                     xmin = rep(NA, length(unique(geodata_timezone$day_EDT))),
                     xmax = rep(NA, length(unique(geodata_timezone$day_EDT))),
                     ymin = rep(-Inf, length(unique(geodata_timezone$day_EDT))),
                     ymax = rep(Inf, length(unique(geodata_timezone$day_EDT))))

for (day in unique(geodata_timezone$day_EDT)){
  if (!day == "23"){
    day_df[paste(day), 2] <- which(geodata_timezone$day_EDT == day)[1]
    day_df[paste(day), 3] <- which(geodata_timezone$day_EDT == as.numeric(as.character(day))+1)[1]-1
  } else {
    day_df[paste(day), 2] <- which(geodata_timezone$day_EDT == day)[1]
    day_df[paste(day), 3] <- nrow(geodata_timezone)
  }
}

ggplot(aes(x = seq_along(ele), y = ele), data = geodata_timezone) +
  geom_rect(data=day_df, inherit.aes=FALSE,
           aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill = factor(group)), col = "grey", alpha=0.5) +
  #annotate("rect", xmin=0, xmax=0, ymin=-Inf, ymax=0, fill="red", alpha=0.5) +
  geom_path() +
  labs(x="Index", y="Elevation (line) and day (rectangles)", title="Elevation per day") +
  scale_y_continuous(limits = c(min(c(geodata_timezone$ele, as.numeric(as.character(geodata_timezone$day_EDT)))), 
                                max(c(geodata_timezone$ele, as.numeric(as.character(geodata_timezone$day_EDT)))))) +
  scale_fill_discrete(name="Days", labels = labels)
```

## Extracting speed data from the gpx files

Looking at the gpx file, there is also speed and course data. It is saved into the extensions tag. However, not all data points seem to have a speed tag.

```{r echo=TRUE, message=FALSE, warning=FALSE, cache = TRUE}
for (i in 1:length(myfiles)){

  tryCatch({
    singleString <- paste(readLines(myfiles[i]), collapse=" ")
    pfile2 <- gsub("</time><extensions><gpxtpx:TrackPointExtension><gpxtpx:course>",
                   "</time><extensions><gpxtpx:TrackPointExtension><gpxtpx:speed>0.00</gpxtpx:speed><gpxtpx:course>", singleString)

    pfile <- htmlTreeParse(pfile2, useInternalNodes = T)

    # Get all elevations, times and coordinates via the respective xpath
    times <- xpathSApply(pfile, path = "//trkpt/time", xmlValue)
    coords <- xpathSApply(pfile, path = "//trkpt", xmlAttrs)
    # Extract latitude and longitude from the coordinates
    lats <- as.numeric(as.character(coords["lat",]))
    lons <- as.numeric(as.character(coords["lon",]))

    speed <- xpathSApply(pfile, path = "//trkpt/extensions/*/speed", xmlValue)# in m/s
    
    # course gives the degrees where the car was headed at that point
    course <- xpathSApply(pfile, path = "//trkpt/extensions/*/course", xmlValue)
    # Put everything in a dataframe and get rid of old variables
    geodf <- data.frame(lat = lats, lon = lons, time = times, speed = speed, course = course)

    if (i == 1){
      geodata_new <- geodf
    } else {
      geodata_new <- rbind(geodata_new, geodf)
    }

    rm(pfile, pfile2, singleString)
  }, error=function(e){cat("ERROR\n")})

}
```

```{r echo=TRUE, message=FALSE, warning=FALSE, cache = TRUE}
# Transforming the time column
geodata_new$time <- strptime(geodata_new$time, format = "%Y-%m-%dT%H:%M:%SZ")

# ordering by date
geodata_new <- arrange(geodata_new, time)

# removing all tracks from Germany
geodata_new_trip <- geodata_new[which(as.numeric(as.character(geodata_new$lon)) < 0),]

# adding a column with day for separating the plotting by day lateron
geodata_new_trip$day <- as.factor(format(as.Date(geodata_new_trip$time,format="%Y-%m-%d"), "%d"))

head(geodata_new_trip)
```

Garmin shows speed as m/s, so I am converting it to km/h.

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height= 5}
library(yarrr)

geodata_new_trip$speed_kmh <- as.numeric(as.character(geodata_new_trip$speed)) *3.6

speed_data <- geodata_new_trip[which(as.numeric(as.character(geodata_new_trip$speed_kmh)) > 0),]
speed_data <- speed_data[!duplicated(speed_data), ]

summary(as.numeric(as.character(speed_data$speed_kmh)))

pirateplot(formula = as.numeric(as.character(speed_kmh)) ~ as.factor(day),
           point.o = .1,
           data = speed_data,
           ylab = "Speed (km/h)",
           xlab = "Day in September",
           main = "Speed per day")
```



